Missing The Point: Non-Convergence in Iterative Imputation Algorithms

### 2021-09-17 Manuscript.Rmd
re-order paragraphs in intro

Iterative imputation has become the de facto standard to accommodate missing data in social scientific research. The technique enables researchers to draw valid inferences despite having incomplete observations. Unbiased, confidence-valid estimates are obtained by separating the missing data problem from the scientific problem. First, the missing values are imputed (i.e., filled in) using some sort of algorithm. Only subsequently, the scientific model of interest is performed on the completed data. To obtain valid scientific estimates, both the missing data problem and the scientific problem should be appropriately considered. The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputed values.

The scientific inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial. There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms. A widely accepted practice is visual inspection of the algorithm, although diagnosing convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. Therefore, a quantitative diagnostic method to assess convergence would be preferred.*

It is challenging to arrive upon a single point at which convergence has been reached.* Since the aim is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged. Because of this property, it may be more desirable to focus on *non*-convergence. Fortunately, there are non-convergence identifiers for other iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.

@@

Add reference: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p. 285).

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence, we terminate the algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios that are defined by the data generating mechanism.

>>>>

Iterative imputation has become the de facto standard to accommodate missing data in social scientific research [ref: e.g., rubin after 18 years paper? or murray 2018?]. The technique enables researchers to draw valid inferences when faced with [phrasing: despite of?] incomplete observations. Every missing cell in an incomplete dataset is imputed (i.e., filled in) using an imputation algorithm. And once the dataset is completed, an analysis of scientific interest can be performed. This analysis will yield unbiased and confidence-valid estimates when both the missing data problem and the scientific problem are appropriately considered. The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputed values. Yet, there has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.

<!-- Any scientific inferences obtained after iterative imputation rely on the convergence of the imputation algorithm,  -->

Determining whether an algorithm has converged is not trivial, especially in the context of iterative imputation.* It is challenging to arrive upon a single point at which convergence has been reached. Since the aim of iterative imputation is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged. Because of this property, it may be more desirable to focus on *non*-convergence. A widely accepted practice is visual inspection of the algorithm [@ragh07; @buur18]. Diagnosing non-convergence through visual inspection may, however, be undesirable: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. [add argument: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p. 285).] Therefore, a quantitative diagnostic method to assess convergence would be preferred. Fortunately, there are non-convergence identifiers for other iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.

@@

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To assess non-convergence, we terminate the algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios that are defined by the data generating mechanism.
